### 2020-08-20 submission/manuscript.tex
Additional edits

The gut microbiota has a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use animals obtained
from a single source. We treated mice from 6 sources (2 University of
Michigan colonies and 4 vendors) with 1 clindamycin dose, followed by
\emph{C. difficile} challenge 1 day later and then measured \emph{C.
difficile} colonization levels through 9 days post-infection. The
microbiota were profiled via 16S rRNA gene sequencing to examine the
variation across sources and alterations due to clindamycin treatment
and \emph{C. difficile} challenge. While all mice were colonized 1-day
post-infection, variation emerged from days 3-7 post-infection with
animals from some sources colonized with \emph{C. difficile} for longer
and at higher levels. We identified bacteria that varied in relative
abundance across sources and throughout the experiment. Some bacteria
were consistently impacted by clindamycin treatment in all sources of
mice including \emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
\emph{Enterobacteriaceae}. To identify bacteria that were most important
to colonization regardless of the source, we created logistic regression
models that successfully classified mice based on whether they cleared
\emph{C. difficile} by 7 days post-infection using baseline,
post-clindamycin, and 1-day post-infection community composition data.
With these models, we identified 4 bacteria that were predictive of
whether \emph{C. difficile} cleared. They varied across sources
(\emph{Bacteroides}), were altered by clindamycin
(\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
\emph{Enterococcus}). Microbiota variation across sources better
emulates human inter-individual variation and can help identify
bacterial drivers of phenotypic variation in the context of CDIs.

@@

Antibiotics are a common risk factor for \emph{Clostridioides difficile}
infections (CDIs) due to there effect on the intestinal microbiota, but
there is variation in who goes on to develop severe or recurrent CDIs
after exposure (1, 2). Additionally, asymptomatic colonization, where
\emph{C. difficile} is detectable, but symptoms are absent, has been
documented in infants and adults (3, 4). The intestinal microbiota has
been implicated in asymptomatic colonization (5, 6), susceptibility to
CDIs (7), and adverse CDI outcomes (9--12). However, it is not clear how
much inter-individual microbiota variation contributes to the range of
outcomes observed after \emph{C. difficile} exposure relative to other
risk factors.

Mouse models of CDIs have been a great tool for understanding \emph{C.
difficile} pathogenesis (13). The number of CDI mouse model studies has
grown substantially since Chen et al.~published their C57BL/6 model in
2008, which disrupted the gut microbiota with antibiotics to enable
\emph{C. difficile} colonization and symptoms such as diarrhea and
weight loss (14). CDI mouse models have been used to examine
translationally relevant questions regarding \emph{C. difficile},
including the role of the microbiota and efficacy of potential
therapeutics for treating CDIs (15). However, variation in the
microbiota between mice from the same breeding colony is much less than
the inter-individual variation observed between humans (16, 17).
Studying CDIs in mice with a homogenous microbiota is likely to
overstate the importance of individual mechanisms. Using mice that have
a more heterogenous microbiota would allow researchers to identify and
validate more generalizable mechanisms responsible for CDI.

In the past, our group has attempted to introduce more variation into
the mouse microbiota by using a variety of antibiotic treatments
(18--21). An alternative approach to maximize microbiota variation is to
use mice from multiple sources (22, 23). The differences between the
microbiota of mice from vendors have been well documented and shown to
influence susceptibility to a variety of diseases (24, 25), including
enteric infections (22, 23, 26--30). Different research groups have also
observed variation in CDI outcomes despite using similar murine models
(13, 18, 21, 31--33). Here we examined how variations in the baseline
microbiota and responses to clindamycin treatment in C57BL/6 mice from
six different sources influenced susceptibility to \emph{C. difficile}
colonization and the time needed to clear the infection.

@@

\textbf{The variation in the microbiota is high between mice from
different sources.} We obtained C57BL/6 mice from 6 different sources:
two colonies from the University of Michigan that were split from each
other in 2010 (the Young and Schloss lab colonies) and four commercial
sources: the Jackson Laboratory, Charles River Laboratories, Taconic
Biosciences, and Envigo (which was formerly Harlan). These 4 vendors
were chosen because they are commonly used for murine CDI studies (26,
34--40). Two experiments were conducted, approximately 3 months apart.

We sequenced the 16S rRNA gene from fecal samples collected from these
mice after they acclimated to the University of Michigan animal housing
environment. We first examined the alpha diversity across the 6 sources
of mice. There was a significant difference in the richness (i.e.~number
of observed operational taxonomic units (OTUs)), but not Shannon
diversity index (\emph{P}\textsubscript{FDR} = 0.03 and
\emph{P}\textsubscript{FDR} = 0.052, respectively) across the sources of
mice (Fig. 1A-B and Tables S1-2). Next, we compared the community
structure of mice (Fig. 1C). The source of mice and the interactions
between the source and cage effects explained most of the observed
variation between fecal communities (PERMANOVA combined
R\textsuperscript{2} = 0.90, \emph{P} \textless{} 0.001, Fig. 1C and
Table S3). Mice that are co-housed tend to have similar gut microbiotas
due to coprophagy (41). Since mice within the same source were housed
together, it was not surprising that the cage effect also contributed to
the observed community variation. There were some differences between
the 2 experiments we conducted, as the experiment and cage effects
significantly explained the observed community variation for the Schloss
and Young lab mouse colonies (Fig. S2A-B and Table S4). However, most of
the vendors also clustered by experiment (Fig. S2C-D, F), suggesting
there was some community variation between the 2 experiments within each
source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H).
After finding differences at the community level, we next identified the
bacteria that varied between sources of mice. There were 268 OTUs with
relative abundances that were significantly different between the
sources (Fig. 1D and Table S5). Though we saw differences between
experiments at the community level, there were no OTUs that were
significantly different between experiments within Schloss, Young, and
Envigo mice (all \emph{P} \textgreater{} 0.05). By using mice from six
sources we were able to increase the variation in the starting
communities to evaluate in a clindamycin-based CDI model.

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization, but clearance time varies across sources.}
Clindamycin is frequently implicated with human CDIs (42) and was part
of the antibiotic treatment for the frequently cited 2008 CDI mouse
model (14). We have previously demonstrated mice are rendered
susceptible to \emph{C. difficile}, but clear the pathogen within 9 days
when treated with clindamycin alone (21, 43). All mice were treated with
10 mg/kg clindamycin via intraperitoneal injection and one day later
challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores
(Fig. 2A). The day after infection, \emph{C. difficile} was detectable
in all mice at a similar level (median CFU range: 2.2e+07-1.3e+08;
\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
mice susceptible regardless of source (Fig. 2B). However, between 3 and
7 days post-infection, we observed variation in \emph{C. difficile}
levels across sources of mice (all \emph{P}\textsubscript{FDR} \(\le\)
0.019; Fig. 2B and Table S6). This suggested the mouse source was
associated with \emph{C. difficile} clearance. While the colonization
dynamics were similar between the two experiments, the Schloss mice took
longer to clear in the 1st experiment compared to the 2nd and the Envigo
mice took longer to clear in the 2nd experiment compared to the 1st
(Fig. S2A-B). The change in the mice's weight significantly varied
across sources of mice with the most weight loss occurring two days
post-infection (Fig. 2C and Table S7). There was also one Jackson and
one Envigo mouse that died between 1- and 3-days post-infection during
the second experiment. Mice obtained from Jackson, Taconic, and Envigo
tended to lose more weight, have higher \emph{C. difficile} CFU levels
and take longer to clear the infection compared to the other sources of
mice (although there was variation between experiments with Schloss and
Envigo mice). This was particularly evident 7 days post-infection (Fig.
2B-C, Fig. S2C-D), when 57\% of the mice were still colonized with
\emph{C. difficile} (Fig. S2E). By 9 days post-infection the majority of
the mice from all sources had cleared \emph{C. difficile} (Fig. 2B) with
the exception of 1 Taconic mouse from the first experiment and 2 Envigo
mice from the second experiment. Thus, clindamycin rendered all mice
susceptible to \emph{C. difficile} 630 colonization, regardless of
source, but there was significant variation in disease phenotype across
the sources of mice.

\textbf{Clindamycin treatment alters bacteria in all sources, but a
subset of bacterial differences across sources persists.} Given the
variation in fecal communities that we observed across breeding
colonies, we hypothesized that variation in \emph{C. difficile}
clearance would be explained by community variation across the 6 sources
of mice. As expected, clindamycin treatment decreased the richness and
Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly,
significant differences in diversity metrics between sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after clindamycin
treatment, with Charles River mice having higher richness and Shannon
diversity than most of the other sources (Fig 3A-B and Tables S1-2). The
clindamycin treatment decreased the variation in community structures
between sources of mice. The source of mice and the interactions between
source and cage effects explained almost all of the observed variation
between communities (combined R\textsuperscript{2} = 0.99, \emph{P}
\textless{} 0.001; Fig. 3C and Table S3). However, there were only 18
OTUs (Fig. 3D and Table S8) with relative abundances that significantly
varied between sources. Next, we identified the bacteria that shifted
after clindamycin treatment, regardless of source by analyzing paired
fecal samples from mice that were collected at baseline and after
clindamycin treatment. We identified 153 OTUs that were altered after
clindamycin treatment in most mice (Fig. 3E and Table S9). When we
compared the list of significant clindamycin impacted bacteria with the
bacteria that varied between sources post-clindamycin, we found 4 OTUs
(\emph{Enterobacteriaceae (OTU 1)}, \emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) overlapped
(Fig. 3D-E and Tables S8-9). Importantly, some of the OTUs that varied
between sources also shifted with clindamycin treatment. For example,
\emph{Proteus} increased after clindamycin treatment (Fig. 3D), but only
in Taconic mice. \emph{Enterococcus} was primarily found only in mice
purchased from commercial vendors and also increased in relative
abundance after clindamycin treatment (Fig. 3D). These findings
demonstrate that clindamycin had a consistent impact on the fecal
bacterial communities of mice from all sources and only a subset of the
OTUs continued to vary between sources.

\textbf{Microbiota variation between sources is maintained after
\emph{C. difficile} challenge.} One day post-infection, significant
differences in diversity metrics remained across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
S1-2). Although the Charles River mice had more diverse communities and
were also able to clear \emph{C. difficile} faster than the other
sources, diversity did not explain the observed variation in \emph{C.
difficile} colonization across sources. The Young and Schloss mice had
the lowest diversity 1 day post-infection and were able to clear
\emph{C. difficile} earlier than Jackson, Taconic and Envigo mice. The
source of mice and the interactions between source and cage effects
continued to explain most of the observed community variation (combined
R\textsuperscript{2} = 0.88; \emph{P} \textless{} 0.001; Fig. 4C and
Table S3). One day after \emph{C. difficile} challenge, there were 44
OTUs (Fig. 4D and Table S10) with significantly different relative
abundances across sources.

Throughout the experiment, the source of mice continued to be the
dominant factor that explained the observed variation across fecal
communities (PERMANOVA R\textsuperscript{2} = 0.35, \emph{P} \textless{}
0.001) followed by interactions between cage effects and the day of the
experiment (Movie S1 and Table S11). Mice fecal samples from the same
source of mice continued to cluster closely to each other throughout the
experiment. By 7 days post-infection, when approximately 43\% mice had
cleared \emph{C. difficile}, most of the mice had not recovered to their
baseline community structure (Fig. 4E). The distance to the baseline
community did not explain the variation in \emph{C. difficile} clearance
as the Schloss and Young mice had mostly cleared \emph{C. difficile},
but their communities were a greater distance from baseline 7 days
post-infection compared to the Jackson and Taconic mice that were still
colonized. In summary, mouse bacterial communities varied significantly
between sources throughout the course of the experiment and a consistent
subset of bacteria remained different between sources regardless of
clindamycin and \emph{C. difficile} challenge.

\textbf{Baseline, post-clindamycin, and post-infection community data
can predict mice that will clear \emph{C. difficile} by 7 days
post-infection.} After identifying taxa that varied between sources,
changed after clindamycin treatment, or both, we determined which taxa
were influencing the variation in \emph{C. difficile} colonization at
day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic
regression models with input bacterial community data from the baseline
(day = -1), post-clindamycin (day = 0), and post-infection (day = 1)
timepoints of the experiment to predict \emph{C. difficile} colonization
status on day 7 (Fig. S3A-B). All models were better at predicting
\emph{C. difficile} colonization status on day 7 than random chance (all
\emph{P} \textless{} 0.001, Table S12). The model based on the
post-clindamycin (AUROC = 0.78) community OTU data performed
significantly better than the baseline (AUROC = 0.72) or the
post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
\textless{} 0.001 for pairwise comparisons, Fig. S3C and Table S13).
Thus, we were able to use bacterial relative abundance data from the
time of \emph{C. difficile} challenge to differentiate mice that had
cleared \emph{C. difficile} before day 7 from the mice still colonized
with \emph{C. difficile} at that timepoint. This result suggests the
bacterial community's response to clindamycin treatment had the greatest
influence on subsequent \emph{C. difficile} colonization dynamics.

To examine the bacteria that were driving each model's performance, we
selected the 20 OTUs that had the highest absolute feature weights in
each of the 3 models (Table S14). First, we looked at OTUs from the
model with the best performance, which was based on the post-clindamycin
treatment (day 0) bacterial community data. Out of the 10 highest ranked
OTUs, 7 OTUs (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
\emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
and \emph{Ruminococcaceae}) were associated with \emph{C. difficile}
colonization 7 days post-infection, while 3 OTUs
(\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
\emph{Porphyromonadaceae}) were associated with clearance (Fig. 5A).
Next, we examined whether any of the top 20 ranked OTUs from the
post-clindamycin (day 0) model were also important in the other 2
classification models based on baseline (day -1) and 1 day
post-infection community data. We identified 6 OTUs
(\emph{Enterobacteriaceae, Ruminococcaceae, Lactobacillus, Bacteroides,
Porphyromonadaceae, Erysipelotrichaceae}) that were important to the
post-clindamycin model and either the baseline or 1 day post-infection
models (Table S14). Thus, a subset of bacterial OTUs were important for
determining \emph{C. difficile} colonization dynamics across multiple
timepoints.

To determine whether the OTUs driving the classification models also
varied between sources, were altered by clindamycin treatment, or both,
we identified the OTUs from each model that varied between sources (Fig.
1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin
treatment (Fig. 3E and Table S9; Fig. S4). Comparing the features
important to the 3 models identified 14 OTUs associated with source, 21
OTUs associated with clindamycin treatment, and 6 OTUs associated with
both (Fig. 5B). Together, these results suggest that the initial
bacterial communities and their responses to clindamycin influence the
clearance of \emph{C. difficile}.

Several OTUs (\emph{Bacteroides, Enterococcus, Enterobacteriaceae,
Porphyromonadaceae}) that overlapped with our previous analyses appeared
across at least 2 models, so we examined how the relative abundances of
these OTUs varied over the course of the experiment (Fig. 6). Across the
9 days post-infection, there was at least 1 timepoint when the relative
abundances of these OTUs significantly varied between sources (Table
S15). Interestingly, there were no OTUs that emerged as consistently
enriched or depleted in mice that were colonized past 7 days
post-infection, suggesting that multiple bacteria influence \emph{C.
difficile} colonization dynamics.

@@

Applying our CDI model with mice from 6 different sources, allowed us to
identify bacterial taxa that were unique to different sources as well as
taxa that were universally impacted by clindamycin. We trained L2
logistic regression models with baseline (day -1), post-clindamycin
treatment (day 0), and 1-day post-infection fecal community data that
could predict whether mice cleared \emph{C. difficile} by 7 days
post-infection better than random chance. We identified
\emph{Bacteroides, Enterococcus, Enterobacteriaceae, Porphyromonadaceae}
(Fig. 6) as candidate bacteria within these communities that were
influencing variation in \emph{C. difficile} colonization dynamics since
these bacteria were all important in the logistic regression models and
varied by source, were impacted by clindamycin treatment, or both.
Overall, our results demonstrated clindamycin was sufficient to render
mice from multiple sources susceptible to CDI and only a subset of the
inter-individual microbiota variation across mice from different sources
was associated with the time needed to clear \emph{C. difficile}.

Other studies have taken similar approaches by using mice from multiple
sources to identify bacteria that either promote colonization resistance
or increase susceptibility to enteric infections (22, 23, 26--30). For
example, against \emph{Salmonella} infections, \emph{Enterobacteriaceae}
and segmented filamentous bacteria have emerged as protective (22, 27).
We found \emph{Enterobacteriaceae} increased in all sources of mice
after clindamycin treatment, facilitating \emph{C. difficile}
colonization. However, there was also variation in
\emph{Enterobacteriaceae} relative abundance levels between sources that
was associated with the variation in \emph{C. difficile} colonization
dynamics across sources. Thus, bacteria may have differential roles in
determining susceptibility depending on the type of bacterial infection.

>>>>

The gut microbiota has a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use animals obtained
from a single source. We treated mice from 6 sources (2 University of
Michigan colonies and 4 commercial vendors) with clindamycin, followed
by a \emph{C. difficile} challenge and then measured \emph{C. difficile}
colonization levels throughout the infection. The microbiota were
profiled via 16S rRNA gene sequencing to examine the variation across
sources and alterations due to clindamycin treatment and \emph{C.
difficile} challenge. While all mice were colonized 1-day
post-infection, variation emerged from days 3-7 post-infection with
animals from some sources colonized with \emph{C. difficile} for longer
and at higher levels. We identified bacteria that varied in relative
abundance across sources and throughout the experiment. Some bacteria
were consistently impacted by clindamycin treatment in all sources of
mice including \emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and
\emph{Enterobacteriaceae}. To identify bacteria that were most important
to colonization regardless of the source, we created logistic regression
models that successfully classified mice based on whether they cleared
\emph{C. difficile} by 7 days post-infection using community composition
data at baseline, post-clindamycin, and 1-day post-infection. With these
models, we identified 4 bacteria that were predictive of whether
\emph{C. difficile} cleared. They varied across sources
(\emph{Bacteroides}), were altered by clindamycin
(\emph{Porphyromonadaceae}), or both (\emph{Enterobacteriaceae} and
\emph{Enterococcus}). Allowing for microbiota variation across sources
better emulates human inter-individual variation and can help identify
bacterial drivers of phenotypic variation in the context of CDIs.

@@

Antibiotics are a common risk factor for \emph{Clostridioides difficile}
infections (CDIs) due to their effect on the intestinal microbiota, but
there is variation in who goes on to develop severe or recurrent CDIs
after exposure (1, 2). Additionally, asymptomatic colonization, where
\emph{C. difficile} is detectable, but symptoms are absent, has been
documented in infants and adults (3, 4). The intestinal microbiota has
been implicated in asymptomatic colonization (5, 6), susceptibility to
CDIs (7), and adverse CDI outcomes (9--12). However, it is not clear how
much inter-individual microbiota variation contributes to the range of
outcomes observed after \emph{C. difficile} exposure relative to other
risk factors.

Mouse models of CDIs have been a great tool for understanding \emph{C.
difficile} pathogenesis (13). The number of CDI mouse model studies has
grown substantially since Chen et al.~published their C57BL/6 model in
2008, which disrupted the gut microbiota with antibiotics to enable
\emph{C. difficile} colonization and symptoms such as diarrhea and
weight loss (14). CDI mouse models have been used to examine
translationally relevant questions regarding \emph{C. difficile},
including the role of the microbiota and efficacy of potential
therapeutics for treating CDIs (15). However, variation in the
microbiota between mice from the same breeding colony is much less than
the inter-individual variation observed between humans (16, 17).
Studying CDIs in mice with a homogeneous microbiota is likely to
overstate the importance of individual mechanisms. Using mice that have
a more heterogeneous microbiota would allow researchers to identify and
validate more generalizable mechanisms responsible for CDI.

In the past, our group has attempted to introduce more variation into
the mouse microbiota by using a variety of antibiotic treatments
(18--21). An alternative approach to maximize microbiota variation is to
use mice from multiple sources (22, 23). The differences between the
microbiota of mice from vendors have been well documented and shown to
influence susceptibility to a variety of diseases (24, 25), including
enteric infections (22, 23, 26--30). Different research groups have also
observed different CDI outcomes despite using similar murine models (13,
18, 21, 31--33). Here we examined how variation in the baseline
microbiota and responses to clindamycin treatment in C57BL/6 mice from
six different sources influenced susceptibility to \emph{C. difficile}
colonization and the time needed to clear the infection.

@@

\textbf{The variation in the microbiota is high between mice from
different sources.} We obtained C57BL/6 mice from 6 different sources:
two colonies from the University of Michigan that were split from each
other in 2010 (the Young and Schloss lab colonies) and four commercial
vendors: the Jackson Laboratory, Charles River Laboratories, Taconic
Biosciences, and Envigo (which was formerly Harlan). These 4 vendors
were chosen because they are commonly used for murine CDI studies (26,
34--40). Two experiments were conducted, approximately 3 months apart.

We sequenced the 16S rRNA gene from fecal samples collected from these
mice after they acclimated to the University of Michigan animal housing
environment. We first examined the alpha diversity across the 6 sources
of mice. There was a significant difference in the richness (i.e.~number
of observed operational taxonomic units (OTUs)), but not Shannon
diversity index across the sources of mice (\emph{P}\textsubscript{FDR}
= 0.03 and \emph{P}\textsubscript{FDR} = 0.052, respectively; Fig. 1A-B
and Tables S1-2). Next, we compared the community structure of mice
(Fig. 1C). The source of mice and the interactions between the source
and cage effects explained most of the observed variation between fecal
communities (PERMANOVA combined R\textsuperscript{2} = 0.90, \emph{P}
\textless{} 0.001; Fig. 1C and Table S3). Mice that are co-housed tend
to have similar gut microbiotas due to coprophagy (41). Since mice
within the same source were housed together, it was not surprising that
the cage effect also contributed to the observed community variation.
There were some differences between the 2 experiments we conducted, as
the experiment and cage effects significantly explained the observed
community variation for the Schloss and Young lab mouse colonies (Fig.
S2A-B and Table S4). However, most of the vendors also clustered by
experiment (Fig. S2C-D, F), suggesting there was some community
variation between the 2 experiments within each source, particularly for
Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences
at the community level, we next identified the bacteria that varied
between sources of mice. There were 268 OTUs with relative abundances
that were significantly different between the sources (Fig. 1D and Table
S5). Though we saw differences between experiments at the community
level, there were no OTUs that were significantly different between
experiments within Schloss, Young, and Envigo mice (all \emph{P}
\textgreater{} 0.05). By using mice from six sources we were able to
increase the variation in the starting communities to evaluate in a
clindamycin-based CDI model.

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization, but clearance time varies across sources.}
Clindamycin is frequently implicated with human CDIs (42) and was part
of the antibiotic treatment for the frequently cited 2008 CDI mouse
model (14). We have previously demonstrated mice are rendered
susceptible to \emph{C. difficile}, but clear the pathogen within 9 days
when treated with clindamycin alone (21, 43). All mice were treated with
10 mg/kg clindamycin via intraperitoneal injection and one day later
challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores
(Fig. 2A). The day after infection, \emph{C. difficile} was detectable
in all mice at a similar level (median CFU range: 2.2e+07-1.3e+08;
\emph{P}\textsubscript{FDR} = 0.15), indicating clindamycin rendered all
mice susceptible regardless of source (Fig. 2B). However, between 3 and
7 days post-infection, we observed variation in \emph{C. difficile}
levels across sources of mice (all \emph{P}\textsubscript{FDR} \(\le\)
0.019; Fig. 2B and Table S6). This suggested the source of mice was
associated with \emph{C. difficile} clearance. While the colonization
dynamics were similar between the two experiments, the Schloss mice took
longer to clear in the first experiment compared to the second and the
Envigo mice took longer to clear in the second experiment compared to
the first (Fig. S2A-B). The change in the mice's weight significantly
varied across sources of mice with the most weight loss occurring two
days post-infection (Fig. 2C and Table S7). There was also one Jackson
and one Envigo mouse that died between 1- and 3-days post-infection
during the second experiment. Mice obtained from Jackson, Taconic, and
Envigo tended to lose more weight, have higher \emph{C. difficile} CFU
levels and take longer to clear the infection compared to the other
sources of mice (although there was variation between experiments with
Schloss and Envigo mice). This was particularly evident 7 days
post-infection (Fig. 2B-C, Fig. S2C-D), when 57\% of the mice were still
colonized with \emph{C. difficile} (Fig. S2E). By 9 days post-infection
the majority of the mice from all sources had cleared \emph{C.
difficile} with the exception of 1 Taconic mouse from the first
experiment and 2 Envigo mice from the second experiment (Fig. 2B). Thus,
clindamycin rendered all mice susceptible to \emph{C. difficile} 630
colonization, regardless of source, but there was significant variation
in disease phenotype across the sources of mice.

\textbf{Clindamycin treatment alters bacteria in all sources, but a
subset of bacterial differences across sources persists.} Given the
variation in fecal communities that we observed across breeding
colonies, we hypothesized that variation in \emph{C. difficile}
clearance would be explained by community variation across the 6 sources
of mice. As expected, clindamycin treatment decreased the richness and
Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly,
significant differences in diversity metrics between sources emerged
after clindamycin treatment, with Charles River mice having higher
richness and Shannon diversity than most of the other sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05; Fig 3A-B and Tables
S1-2). The clindamycin treatment decreased the variation in community
structures between sources of mice. The source of mice and the
interactions between source and cage effects explained almost all of the
observed variation between communities (combined R\textsuperscript{2} =
0.99, \emph{P} \textless{} 0.001; Fig. 3C and Table S3). However, there
were only 18 OTUs with relative abundances that significantly varied
between sources (Fig. 3D and Table S8). Next, we identified the bacteria
that shifted after clindamycin treatment, regardless of source by
analyzing paired fecal samples from mice that were collected at baseline
and after clindamycin treatment. We identified 153 OTUs that were
altered after clindamycin treatment in most mice (Fig. 3E and Table S9).
When we compared the list of significant clindamycin impacted bacteria
with the bacteria that varied between sources post-clindamycin, we found
4 OTUs that were shared between the lists (\emph{Enterobacteriaceae (OTU
1)}, \emph{Lachnospiraceae} (OTU 130), \emph{Lactobacillus} (OTU 6),
\emph{Enterococcus} (OTU 23); Fig. 3D-E and Tables S8-9). Importantly,
some of the OTUs that varied between sources also shifted with
clindamycin treatment. For example, \emph{Proteus} increased after
clindamycin treatment (Fig. 3D), but only in Taconic mice.
\emph{Enterococcus} was primarily found only in mice purchased from
commercial vendors and also increased in relative abundance after
clindamycin treatment (Fig. 3D). These findings demonstrate that
clindamycin had a consistent impact on the fecal bacterial communities
of mice from all sources and only a subset of the OTUs continued to vary
between sources.

\textbf{Microbiota variation between sources is maintained after
\emph{C. difficile} challenge.} One day post-infection, significant
differences in diversity metrics remained across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05, Fig 4A-B and Tables
S1-2). Although the Charles River mice had more diverse communities and
were also able to clear \emph{C. difficile} faster than the other
sources, diversity did not explain the observed variation in \emph{C.
difficile} colonization across sources. The Young and Schloss mice had
the lowest diversity 1 day post-infection and were able to clear
\emph{C. difficile} earlier than Jackson, Taconic and Envigo mice. The
source of mice and the interactions between source and cage effects
continued to explain most of the observed community variation (combined
R\textsuperscript{2} = 0.88; \emph{P} \textless{} 0.001; Fig. 4C and
Table S3). One day after \emph{C. difficile} challenge, there were 44
OTUs with significantly different relative abundances across sources
(Fig. 4D and Table S10).

Throughout the experiment, the source of mice continued to be the
dominant factor that explained the observed variation across fecal
communities (PERMANOVA R\textsuperscript{2} = 0.35, \emph{P} \textless{}
0.001) followed by interactions between cage effects and the day of the
experiment (Movie S1 and Table S11). Fecal samples from the same source
of mice continued to cluster closely to each other throughout the
experiment. By 7 days post-infection, when approximately 43\% mice had
cleared \emph{C. difficile}, most of the mice had not recovered to their
baseline community structure (Fig. 4E). The distance to the baseline
community did not explain the variation in \emph{C. difficile} clearance
as the Schloss and Young mice had mostly cleared \emph{C. difficile},
but their communities were a greater distance from baseline 7 days
post-infection compared to the Jackson and Taconic mice that were still
colonized. In summary, mouse bacterial communities varied significantly
between sources throughout the course of the experiment and a consistent
subset of bacteria remained different between sources regardless of
clindamycin and \emph{C. difficile} challenge.

\textbf{Baseline, post-clindamycin, and post-infection community data
can predict mice that will clear \emph{C. difficile} by 7 days
post-infection.} After identifying taxa that varied between sources,
changed after clindamycin treatment, or both, we determined which taxa
were influencing the variation in \emph{C. difficile} colonization at
day 7 (Fig. 2B, Fig. S2C). We trained three L2-regularized logistic
regression models with either input bacterial community data from the
baseline (day = -1), post-clindamycin (day = 0), or post-infection (day
= 1) timepoints of the experiment to predict \emph{C. difficile}
colonization status on day 7 (Fig. S3A-B). All models were better at
predicting \emph{C. difficile} colonization status on day 7 than random
chance (all \emph{P} \textless{} 0.001, Table S12). The model based on
the post-clindamycin (AUROC = 0.78) community OTU data performed
significantly better than the baseline (AUROC = 0.72) or the
post-infection (AUROC = 0.67) models (\emph{P}\textsubscript{FDR}
\textless{} 0.001 for pairwise comparisons; Fig. S3C and Table S13).
Thus, we were able to use bacterial relative abundance data from the
time of \emph{C. difficile} challenge to differentiate mice that had
cleared \emph{C. difficile} before day 7 from the mice still colonized
with \emph{C. difficile} at that timepoint. This result suggests that
the bacterial community's response to clindamycin treatment had the
greatest influence on subsequent \emph{C. difficile} colonization
dynamics.

To examine the bacteria that were driving each model's performance, we
selected the 20 OTUs that had the highest absolute feature weights in
each of the 3 models (Table S14). First, we looked at OTUs from the
model with the best performance, which was based on the post-clindamycin
treatment (day 0) bacterial community data. Out of the 10 highest ranked
OTUs, 7 OTUs were associated with \emph{C. difficile} colonization 7
days post-infection (\emph{Bacteroides}, \emph{Escherichia/Shigella}, 2
\emph{Lachnospiraceae}, \emph{Lactobacillus}, \emph{Porphyromonadaceae},
and \emph{Ruminococcaceae}), while 3 OTUs were associated with clearance
(\emph{Enterobacteriaceae}, \emph{Lachnospiraceae},
\emph{Porphyromonadaceae}; Fig. 5A). Next, we examined whether any of
the top 20 ranked OTUs from the post-clindamycin (day 0) model were also
important in the other 2 classification models based on baseline (day
-1) and 1 day post-infection community data. We identified 6 OTUs that
were important to the post-clindamycin model and either the baseline or
1 day post-infection models (\emph{Enterobacteriaceae, Ruminococcaceae,
Lactobacillus, Bacteroides, Porphyromonadaceae, Erysipelotrichaceae};
Table S14). Thus, a subset of bacterial OTUs were important for
determining \emph{C. difficile} colonization dynamics across multiple
timepoints.

To determine whether the OTUs driving the classification models also
varied between sources, were altered by clindamycin treatment, or both,
we identified the OTUs from each model that varied between sources (Fig.
1D, 3D, 4D and Tables S5, S8, S10) or were impacted by clindamycin
treatment (Fig. 3E and Table S9; Fig. S4). Comparing the features
important to the 3 models identified 14 OTUs associated with source, 21
OTUs associated with clindamycin treatment, and 6 OTUs associated with
both (Fig. 5B). Together, these results suggest that the initial
bacterial communities and their responses to clindamycin influenced the
clearance of \emph{C. difficile}.

Several OTUs that overlapped with our previous analyses appeared across
at least 2 models (\emph{Bacteroides, Enterococcus, Enterobacteriaceae,
Porphyromonadaceae}), so we examined how the relative abundances of
these OTUs varied over the course of the experiment (Fig. 6). Across the
9 days post-infection, there was at least 1 timepoint when the relative
abundances of these OTUs significantly varied between sources (Table
S15). Interestingly, there were no OTUs that emerged as consistently
enriched or depleted in mice that were colonized past 7 days
post-infection, suggesting that multiple bacteria influence \emph{C.
difficile} colonization dynamics.

@@

Applying our CDI model to 6 different sources of mice, allowed us to
identify bacterial taxa that were unique to different sources as well as
taxa that were universally impacted by clindamycin. We trained logistic
regression models with baseline (day -1), post-clindamycin treatment
(day 0), and 1-day post-infection fecal community data that could
predict whether mice cleared \emph{C. difficile} by 7 days
post-infection better than random chance. We identified
\emph{Bacteroides, Enterococcus, Enterobacteriaceae, Porphyromonadaceae}
(Fig. 6) as candidate bacteria within these communities that influenced
variation in \emph{C. difficile} colonization dynamics since these
bacteria were all important in the logistic regression models and varied
by source, were impacted by clindamycin treatment, or both. Overall, our
results demonstrated clindamycin was sufficient to render mice from
multiple sources susceptible to CDI and only a subset of the
inter-individual microbiota variation across mice from different sources
was needed to predict which mice could clear \emph{C. difficile}.

Other studies have used mice from multiple sources to identify bacteria
that either promote colonization resistance or increase susceptibility
to enteric infections (22, 23, 26--30). For example, against
\emph{Salmonella} infections, \emph{Enterobacteriaceae} and segmented
filamentous bacteria have emerged as protective (22, 27). We found
\emph{Enterobacteriaceae} increased in all sources of mice after
clindamycin treatment, facilitating \emph{C. difficile} colonization.
However, there was also variation in \emph{Enterobacteriaceae} relative
abundance levels between sources that was associated with the variation
in \emph{C. difficile} colonization dynamics across sources. Thus,
bacteria may have differential roles in determining susceptibility
depending on the type of bacterial infection.
